---
title: "Project: Practical Machine Learning Project"
output: html_document
---
*Fazli Haleem*
*April 21, 2020*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

This document presents the final project from Coursera’s course Practical Machine Learning, as part of the Specialization in Data Science. RStudio is used to accomplish this project via knitr package.This analysis aim to be the basis for the course quiz and a prediction assignment writeup. This project aim to predict the way in which 6 participants performed some exercise as described below in the class variable in the training data set.The machine learning algorithm is applied to the 20 test cases available in the test data.

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

Read more: <http://groupware.les.inf.puc-rio.br/har#ixzz3xsbS5bVX>

### Loading Required packages

```{r}
rm(list=ls()) 
setwd("C:/Assignment")
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12345)
```


### Data processing and Cleaning

```{r}
#Seting URL to download the data
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

## download the datasets
training <- read.csv(url(UrlTrain))
testing  <- read.csv(url(UrlTest))

# Create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

The generated datasets both have 160 variables. Those variables have many missing values (NA), that we will delete via the procedure mentioned below.Further, we will also remove the ID variables and those with near Zero variance (NZV) variables.

```{r}
# Removing variables with zero variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
dim(TrainSet)
```

```{r}
# removing variables that are mostly missing (NA)
AllNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet  <- TestSet[, AllNA==FALSE]
dim(TrainSet)
```

Next we remove all those variables which are identification number ranging from 1 to 5 columns

```{r}
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
```

## Correlation Analysis
As a preliminary task it is important to do correlation analysis between the variables.The correlation matrix shows which variables are strongly or weakly related and also tells the direction of the assocaiton among variables. We do the correlation among variables as under:
```{r}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

Variables which are strongly correlated are shown by the colour in the above graph.Principal component analysis (PCA) is appropriate to reduce the number of highly correlated variables into few variables. As shown by the graph,in the majority of cases, correlations between variables are not so high, we think we should not use the technique of PCA here.

## Constructing Prediction Model
We will apply three methods including Random Forests, Decision Tree and Generalized Boosted Model to model the regressions and the best one will be used for the quiz predictions as a prerequisite of passing this course. We will also plot Confusion Matrix at the end of each analysis to better visualize the accuracy of the models.

### Random Forest
Model fit
```{r}
set.seed(12345)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
                          trControl=controlRF)
modFitRandForest$finalModel

```

Making prediction on test data:
```{r}
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatRandForest
```

Now we plot the above results:

```{r}
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest - Accuracy =",
                  round(confMatRandForest$overall['Accuracy'], 4)))

```

## Decision Tree
The decision tree is made as under:
```{r}
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)
```

prediction on Test dataset and Confustion matrix
```{r}
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree
```

Plotting the above prediction and to know how much accurate the results are
```{r}
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 4)))
```

## Generalized Boosted Model
Model fit
```{r}
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```
Prediction on the test dataset
```{r}
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatGBM
```
Plotting the matrix results
```{r}
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))
```

## Applying the Selected Model to the Test Data
The accuracy of the 3 regression modeling methods above are:

1.Random Forest : 0.9963
2.Decision Tree : 0.7368
3.GBM : 0.9839
As the above results show that Random forest model is the most accurate method for prediction.Therefore, we apply the Random Forest model to predict the 20 quiz results (testing dataset) as shown below.

```{r}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```

